---
title: "Winter Flounder Catch Rates"
subtitle: "Fyke Survey Annual Report"
author: "B. P. Galligan, R. G. Balouskus, and M. C. McManus"
format: pdf
editor: visual
date: today
---

```{r}
#| echo: false
#| output: false

## This code chunk loads and cleans the data for analysis

# Load packages
library(ggpubr)
library(iml)
library(lubridate)
library(osmdata)
library(randomForest)
library(readxl)
library(sf)
library(tidyverse)

# Import data
fyke <-  read_excel("data/model-deployment/FykeSets_2024.xlsx")

# Calculate Julian day
fyke$Haul_Date <- as.POSIXlt(fyke$Haul_Date, format="%d-%m-%y")
fyke$Haul_Date_Jul <- as.numeric(format(fyke$Haul_Date, "%j"))
  
# Convert Julian Day to be 1 at Nov 1 as that is earliest conceivable start date
fyke$Haul_Date_Jul <- ifelse(fyke$Haul_Date_Jul < 200,
  fyke$Haul_Date_Jul + 61, fyke$Haul_Date_Jul - 304)

# Add water temp column
fyke$Water_Temp_C <- NA

# Fill in water temp column
for (i in 1:nrow(fyke)){
  
  # If there is a haul and set temp available, use average
  if (!is.na(fyke$Set_Water_Temp[i]) & !is.na(fyke$Haul_Water_Temp[i])){
    fyke$Water_Temp_C[i] <- mean(c(fyke$Set_Water_Temp[i], fyke$Haul_Water_Temp[i]))
  } else {
  
    # If only one temp is available, use that
    if (!is.na(fyke$Set_Water_Temp[i])){
      fyke$Water_Temp_C[i] <- fyke$Set_Water_Temp[i]
    } else if (!is.na(fyke$Haul_Water_Temp[i])){
      fyke$Water_Temp_C[i] <- fyke$Haul_Water_Temp[i]
    }
  }
}

# Select desired columns
fyke <- fyke %>% select(
  Station_ID,
  Pond_ID,
  Haul_Year_Winter,
  Haul_Date_Jul,
  Soak_Time,
  Set_Occ_Year,
  Water_Temp_C,
  WFL_Caught,
  WFL_Frequency)

# Subset for complete cases only
fyke <- fyke %>% drop_na()

# Make all columns numeric or factor
fyke$Station_ID <- as.factor(fyke$Station_ID)
fyke$Pond_ID <- as.factor(fyke$Pond_ID)
fyke$WFL_Caught <- as.factor(fyke$WFL_Caught)

```

This report highlights data from survey year:

```{r}
#| echo: false
print(max(fyke$Haul_Year_Winter))
```

## Catch rates

### Combined index

@fig-combined-index provides an overview of the combined abundance patterns in all three coastal ponds.

```{r}
#| echo: false
#| label: fig-combined-index
#| fig-cap: Fishery-independent abundance index for Rhode Island Winter Flounder in all three coastal ponds. The blue line is a partial dependence plot of Winter Flounder abundance on year.
#| fig-pos: h

## This code chunk calculates an abundance index for all ponds

## Run RFM

# Data frame for RFM
df <- fyke %>% select(-WFL_Caught)

# Bootstrap sample for model training
set.seed(12345)
df_boot <- slice_sample(df, prop = 1, replace = TRUE)

# Out of bag sample for testing
df_oob <- setdiff(df, df_boot)

# Data for RFM
x <- select(df_boot, -WFL_Frequency)
y <- df_boot$WFL_Frequency
xtest <- select(df_oob, -WFL_Frequency)
ytest <- df_oob$WFL_Frequency

# Random forest model
rf <- randomForest(x = x, y = y, xtest = xtest, ytest = ytest,
  ntree = 200, mtry = 4,
  importance = TRUE, proximity = TRUE, keep.forest = TRUE)


## PDP Abundance Index

# Make haul winter an integer
fyke$Haul_Year_Winter <- as.integer(fyke$Haul_Year_Winter)

# Calculate partial dependence
pd <- bind_rows(partialPlot(rf, fyke, x.var = "Haul_Year_Winter",
  n.pt = length(unique(fyke$Haul_Year_Winter)), plot = FALSE))

# Create custom rug to show all sampling by date
rug <- fyke %>%
  select(Haul_Year_Winter, Haul_Date_Jul)

# Prepare empty vector for date
rug$Haul_Date <- NA

# Convert Julian date to date object
for (i in 1:nrow(rug)){
  rug$Haul_Date[i] <- as.Date(rug$Haul_Date_Jul[i],
    origin = paste0(rug$Haul_Year_Winter[i] - 1, "-11-01"))
}

# Make Haul Date a date object
rug$Haul_Date <- as.Date(rug$Haul_Date, format = "%Y-%m-%d")

# Convert Haul Date to decimal years
rug$Haul_Date <- as.numeric(
  year(rug$Haul_Date) + (yday(rug$Haul_Date) - 1) / 
    ifelse(leap_year(rug$Haul_Date), 366, 365))

# Plot index
ggplot(pd, aes(x = x, y = y)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_rug(data = rug, aes(x = Haul_Date), inherit.aes = FALSE,
    sides = "b", alpha = 0.1) +
  xlab("Year") +
  ylab("Abundance (no. per haul)") +
  ggtitle("Winter Flounder Abundance", "RI Coastal Ponds") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

### Point Judith Pond

This section will introduce @fig-pj-index.

```{r}
#| echo: false
#| label: fig-pj-index
#| fig-cap: Fishery-independent abundance index for Winter Flounder in Point Judith Pond. The blue line is a partial dependence plot of Winter Flounder abundance on year.
#| fig-pos: h

## This code chunk calculates an abundance index for all ponds

## Run RFM

# Data frame for RFM
df <- fyke %>% select(-WFL_Caught)

# Subset for point judith pond only
df <- df %>% filter(Pond_ID == "PJ")

# Remove pond column
df <- df %>% select(-Pond_ID)

# Bootstrap sample for model training
set.seed(12345)
df_boot <- slice_sample(df, prop = 1, replace = TRUE)

# Out of bag sample for testing
df_oob <- setdiff(df, df_boot)

# Data for RFM
x <- select(df_boot, -WFL_Frequency)
y <- df_boot$WFL_Frequency
xtest <- select(df_oob, -WFL_Frequency)
ytest <- df_oob$WFL_Frequency

# Random forest model
rf <- randomForest(x = x, y = y, xtest = xtest, ytest = ytest,
  ntree = 200, mtry = 4,
  importance = TRUE, proximity = TRUE, keep.forest = TRUE)


## PDP Abundance Index

# Make haul winter an integer
df$Haul_Year_Winter <- as.integer(df$Haul_Year_Winter)

# Calculate partial dependence
pd <- bind_rows(partialPlot(rf, df, x.var = "Haul_Year_Winter",
  n.pt = length(unique(df$Haul_Year_Winter)), plot = FALSE))

# Create custom rug to show all sampling by date
rug <- df %>%
  select(Haul_Year_Winter, Haul_Date_Jul)

# Prepare empty vector for date
rug$Haul_Date <- NA

# Convert Julian date to date object
for (i in 1:nrow(rug)){
  rug$Haul_Date[i] <- as.Date(rug$Haul_Date_Jul[i],
    origin = paste0(rug$Haul_Year_Winter[i] - 1, "-11-01"))
}

# Make Haul Date a date object
rug$Haul_Date <- as.Date(rug$Haul_Date, format = "%Y-%m-%d")

# Convert Haul Date to decimal years
rug$Haul_Date <- as.numeric(
  year(rug$Haul_Date) + (yday(rug$Haul_Date) - 1) / 
    ifelse(leap_year(rug$Haul_Date), 366, 365))

# Plot index
ggplot(pd, aes(x = x, y = y)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_rug(data = rug, aes(x = Haul_Date), inherit.aes = FALSE,
    sides = "b", alpha = 0.1) +
  xlab("Year") +
  ylab("Abundance (no. per haul)") +
  ggtitle("Winter Flounder Abundance", "Point Judith Pond") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

### Potter Pond

This section will introduce @fig-pp-index.

```{r}
#| echo: false
#| label: fig-pp-index
#| fig-cap: Fishery-independent abundance index for Winter Flounder in Potter Pond. The blue line is a partial dependence plot of Winter Flounder abundance on year.
#| fig-pos: h

## This code chunk calculates an abundance index for all ponds

## Run RFM

# Data frame for RFM
df <- fyke %>% select(-WFL_Caught)

# Subset for point judith pond only
df <- df %>% filter(Pond_ID == "PP")

# Remove pond column
df <- df %>% select(-Pond_ID)

# Subset for post 2010
df <- df %>% filter(Haul_Year_Winter > 2009)

# Bootstrap sample for model training
set.seed(12345)
df_boot <- slice_sample(df, prop = 1, replace = TRUE)

# Out of bag sample for testing
df_oob <- setdiff(df, df_boot)

# Data for RFM
x <- select(df_boot, -WFL_Frequency)
y <- df_boot$WFL_Frequency
xtest <- select(df_oob, -WFL_Frequency)
ytest <- df_oob$WFL_Frequency

# Random forest model
rf <- randomForest(x = x, y = y, xtest = xtest, ytest = ytest,
  ntree = 200, mtry = 4,
  importance = TRUE, proximity = TRUE, keep.forest = TRUE)


## PDP Abundance Index

# Make haul winter an integer
df$Haul_Year_Winter <- as.integer(df$Haul_Year_Winter)

# Calculate partial dependence
pd <- bind_rows(partialPlot(rf, df, x.var = "Haul_Year_Winter",
  n.pt = length(unique(df$Haul_Year_Winter)), plot = FALSE))

# Create custom rug to show all sampling by date
rug <- df %>%
  select(Haul_Year_Winter, Haul_Date_Jul)

# Prepare empty vector for date
rug$Haul_Date <- NA

# Convert Julian date to date object
for (i in 1:nrow(rug)){
  rug$Haul_Date[i] <- as.Date(rug$Haul_Date_Jul[i],
    origin = paste0(rug$Haul_Year_Winter[i] - 1, "-11-01"))
}

# Make Haul Date a date object
rug$Haul_Date <- as.Date(rug$Haul_Date, format = "%Y-%m-%d")

# Convert Haul Date to decimal years
rug$Haul_Date <- as.numeric(
  year(rug$Haul_Date) + (yday(rug$Haul_Date) - 1) / 
    ifelse(leap_year(rug$Haul_Date), 366, 365))

# Plot index
ggplot(pd, aes(x = x, y = y)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_rug(data = rug, aes(x = Haul_Date), inherit.aes = FALSE,
    sides = "b", alpha = 0.1) +
  xlab("Year") +
  ylab("Abundance (no. per haul)") +
  ggtitle("Winter Flounder Abundance", "Potter Pond") +
  coord_cartesian(ylim = c(0, 5)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

### Ninigret Pond

This section will introduce @fig-np-index.

```{r}
#| echo: false
#| label: fig-np-index
#| fig-cap: Fishery-independent abundance index for Winter Flounder in Ninigret Pond. The blue line is a partial dependence plot of Winter Flounder abundance on year.
#| fig-pos: h

## This code chunk calculates an abundance index for all ponds

## Run RFM

# Data frame for RFM
df <- fyke %>% select(-WFL_Caught)

# Subset for point judith pond only
df <- df %>% filter(Pond_ID == "NP")

# Remove pond column
df <- df %>% select(-Pond_ID)

# Bootstrap sample for model training
set.seed(12345)
df_boot <- slice_sample(df, prop = 1, replace = TRUE)

# Out of bag sample for testing
df_oob <- setdiff(df, df_boot)

# Data for RFM
x <- select(df_boot, -WFL_Frequency)
y <- df_boot$WFL_Frequency
xtest <- select(df_oob, -WFL_Frequency)
ytest <- df_oob$WFL_Frequency

# Random forest model
rf <- randomForest(x = x, y = y, xtest = xtest, ytest = ytest,
  ntree = 200, mtry = 4,
  importance = TRUE, proximity = TRUE, keep.forest = TRUE)


## PDP Abundance Index

# Make haul winter an integer
df$Haul_Year_Winter <- as.integer(df$Haul_Year_Winter)

# Calculate partial dependence
pd <- bind_rows(partialPlot(rf, df, x.var = "Haul_Year_Winter",
  n.pt = length(unique(df$Haul_Year_Winter)), plot = FALSE))

# Create custom rug to show all sampling by date
rug <- df %>%
  select(Haul_Year_Winter, Haul_Date_Jul)

# Prepare empty vector for date
rug$Haul_Date <- NA

# Convert Julian date to date object
for (i in 1:nrow(rug)){
  rug$Haul_Date[i] <- as.Date(rug$Haul_Date_Jul[i],
    origin = paste0(rug$Haul_Year_Winter[i] - 1, "-11-01"))
}

# Make Haul Date a date object
rug$Haul_Date <- as.Date(rug$Haul_Date, format = "%Y-%m-%d")

# Convert Haul Date to decimal years
rug$Haul_Date <- as.numeric(
  year(rug$Haul_Date) + (yday(rug$Haul_Date) - 1) / 
    ifelse(leap_year(rug$Haul_Date), 366, 365))

# Plot index
ggplot(pd, aes(x = x, y = y)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_rug(data = rug, aes(x = Haul_Date), inherit.aes = FALSE,
    sides = "b", alpha = 0.1) +
  xlab("Year") +
  ylab("Abundance (no. per haul)") +
  ggtitle("Winter Flounder Abundance", "Ninigret Pond") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

## Water temp

This section will introduce @fig-water-temp.

```{r}
#| echo: false
#| label: fig-water-temp
#| fig-cap: Partial dependence plot of Winter Flounder abundance on water temperature. The blue line is the partial dependence plot, and the gray rug shows the distribution of water temperatures in survey data. Scatter plot points represent the observations made in the most recent survey year.
#| fig-pos: h

## This code chunk calculates PDP for water temperature and plots most recent year's data

## Run RFM

# Data frame for RFM
df <- fyke %>% select(-WFL_Caught)

# Bootstrap sample for model training
set.seed(12345)
df_boot <- slice_sample(df, prop = 1, replace = TRUE)

# Out of bag sample for testing
df_oob <- setdiff(df, df_boot)

# Data for RFM
x <- select(df_boot, -WFL_Frequency)
y <- df_boot$WFL_Frequency
xtest <- select(df_oob, -WFL_Frequency)
ytest <- df_oob$WFL_Frequency

# Set Seed
set.seed(12345)

# Random forest model
rf <- randomForest(x = x, y = y, xtest = xtest, ytest = ytest,
  ntree = 200, mtry = 4,
  importance = TRUE, proximity = TRUE, keep.forest = TRUE)


## PDP Water Temp

# Extract rug data
rug <- select(df, Water_Temp_C)

# Extract most recent year's data
df2 <- df %>% filter(Haul_Year_Winter == max(Haul_Year_Winter))

# Calculate partial dependence
pd <- bind_rows(partialPlot(rf, df, x.var = "Water_Temp_C",
  n.pt = 50, plot = FALSE))

# Plot PD with rug
ggplot(df2, aes(x = Water_Temp_C, y = WFL_Frequency)) +
  geom_point(alpha = 0.5) +
  geom_line(data = pd, aes(x = x, y = y),
    color = "blue", linewidth = 1, inherit.aes = FALSE) +
  geom_rug(data = rug, aes(x = Water_Temp_C), inherit.aes = FALSE,
    sides = "b", alpha = 0.1) +
  xlab("Water Temp (째C)") +
  ylab("Abundance (no. per haul)") +
  ggtitle("Temperature Effect", "RI Coastal Ponds") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

Descriptive stats comparing most recent year's temp in each pond to mean temp for all years prior to 2020.

This section will introduce @fig-water-temp-pj.

```{r}
#| echo: false
#| label: fig-water-temp-pj
#| fig-cap: Boxplot of historical water temperature observations in Point Judith Pond. The dashed red line represents the mean water temperature for all years prior to 2020.
#| fig-pos: h

## This code chunk plots descriptive stats for water temperature

# Point Judith
df <- fyke %>% filter(Pond_ID == "PJ")

ggplot(df, aes(group = Haul_Year_Winter, x = Haul_Year_Winter, y = Water_Temp_C)) +
  geom_boxplot() +
  geom_hline(yintercept = mean(df$Water_Temp_C[df$Haul_Year_Winter < 2020]),
    linetype = "dashed", color = "red") +
  xlab("Year") +
  ylab("Water Temp (째C)") +
  ggtitle("Water Temperature", "Point Judith Pond") +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

This section will introduce @fig-water-temp-pp.

```{r}
#| echo: false
#| label: fig-water-temp-pp
#| fig-cap: Boxplot of historical water temperature observations in Potter Pond. The dashed red line represents the mean water temperature for all years prior to 2020.
#| fig-pos: h

## This code chunk plots descriptive stats for water temperature

# Potter Pond
df <- fyke %>% filter(Pond_ID == "PP")

ggplot(df, aes(group = Haul_Year_Winter, x = Haul_Year_Winter, y = Water_Temp_C)) +
  geom_boxplot() +
  geom_hline(yintercept = mean(df$Water_Temp_C[df$Haul_Year_Winter < 2020]),
    linetype = "dashed", color = "red") +
  xlab("Year") +
  ylab("Water Temp (째C)") +
  ggtitle("Water Temperature", "Potter Pond") +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

```

This section will introduce @fig-water-temp-np.

```{r}
#| echo: false
#| label: fig-water-temp-np
#| fig-cap: Boxplot of historical water temperature observations in Ninigret Pond. The dashed red line represents the mean water temperature for all years prior to 2020.
#| fig-pos: h

## This code chunk plots descriptive stats for water temperature

# Ninigret Pond
df <- fyke %>% filter(Pond_ID == "NP")

ggplot(df, aes(group = Haul_Year_Winter, x = Haul_Year_Winter, y = Water_Temp_C)) +
  geom_boxplot() +
  geom_hline(yintercept = mean(df$Water_Temp_C[df$Haul_Year_Winter < 2020]),
    linetype = "dashed", color = "red") +
  xlab("Year") +
  ylab("Water Temp (째C)") +
  ggtitle("Water Temperature", "Ninigret Pond") +
  theme_pubr() +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))


```

## Station

Compare ALE for last five years with ALE for everything before that. This uses the classification model, i.e. catch probability and not abundance.

```{r}
#| echo: false

## This code chunk calculates accumulated local effects (ALE) of station on catch probability. No plots are created.

## Run RFM

# Data frame for RFM
df1 <- fyke %>%
  select(-WFL_Frequency) %>%
  filter(Haul_Year_Winter < 2020)
df2 <- fyke %>%
  select(-WFL_Frequency) %>%
  filter(Haul_Year_Winter >= max(Haul_Year_Winter) - 4)

# Bootstrap sample for model training
set.seed(12345)
df1_boot <- slice_sample(df1, prop = 1, replace = TRUE)
set.seed(12345)
df2_boot <- slice_sample(df2, prop = 1, replace = TRUE)

# Out of bag sample for testing
df1_oob <- setdiff(df1, df1_boot)
df2_oob <- setdiff(df2, df2_boot)

# Data for RFM
x1 <- select(df1_boot, -WFL_Caught)
x2 <- select(df2_boot, -WFL_Caught)
y1 <- df1_boot$WFL_Caught
y2 <- df2_boot$WFL_Caught
xtest1 <- select(df1_oob, -WFL_Caught)
xtest2 <- select(df2_oob, -WFL_Caught)
ytest1 <- df1_oob$WFL_Caught
ytest2 <- df2_oob$WFL_Caught

# Set Seed
set.seed(12345)

# Baseline RFM
rf1 <- randomForest(x = x1, y = y1, xtest = xtest1, ytest = ytest1,
  ntree = 200, mtry = 5,
  importance = TRUE, proximity = TRUE, keep.forest = TRUE)

# Set Seed
set.seed(12345)

# Last Five Years RFM
rf2 <- randomForest(x = x2, y = y2, xtest = xtest2, ytest = ytest2,
  ntree = 200, mtry = 5,
  importance = TRUE, proximity = TRUE, keep.forest = TRUE)


## Calculate ALE

# Predictor object to hold model and data
predictor1 <- Predictor$new(rf1, data = x1, y = y1)
predictor2 <- Predictor$new(rf2, data = x2, y = y2)

# Calculate all effects using ALE
effs1 <- FeatureEffects$new(predictor1, grid.size = 50)
effs2 <- FeatureEffects$new(predictor2, grid.size = 50)

# Extract rug data
rug1 <- df1 %>% select(Station_ID)
rug2 <- df2 %>% select(Station_ID)

# Summarize rug as total number of samples per station
rug1 <- rug1 %>% group_by(Station_ID) %>% summarise(n = n())
rug2 <- rug2 %>% group_by(Station_ID) %>% summarise(n = n())

# Extract rf effects for station
ale1 <- effs1$results$Station_ID
ale2 <- effs2$results$Station_ID

# Limit to class = TRUE
ale1 <- filter(ale1, .class == "TRUE")
ale2 <- filter(ale2, .class == "TRUE")

# Select columns
ale1 <- select(ale1, .value, .borders)
ale2 <- select(ale2, .value, .borders)

# Rename columns
colnames(ale1) <- c("Probability", "Station_ID")
colnames(ale2) <- c("Probability", "Station_ID")

# Add sample size to station_effects
ale1 <- left_join(ale1, rug1, by = "Station_ID")
ale2 <- left_join(ale2, rug2, by = "Station_ID")

# Remove rows with no sample size
ale1 <- filter(ale1, !is.na(n))
ale2 <- filter(ale2, !is.na(n))


## Map Station Effects

# Import lat/lon
stations <- read_excel("data/raw-data/FykeStations.xlsx")

# Select columns
stations <- select(stations, Station_ID, Latitude, Longitude)

# Merge with station data
ale1 <- left_join(ale1, stations, by = "Station_ID")
ale2 <- left_join(ale2, stations, by = "Station_ID")

# Group station effects by positive and negative
ale1$posneg <- ifelse(ale1$Probability > 0, "Positive", "Negative")
ale2$posneg <- ifelse(ale2$Probability > 0, "Positive", "Negative")

# Add a pond column
ale1$Pond <- substr(ale1$Station_ID, 1, 2)
ale2$Pond <- substr(ale2$Station_ID, 1, 2)

# Remove rows with NA
ale1 <- filter(ale1, !is.na(Latitude))
ale2 <- filter(ale2, !is.na(Latitude))

# Split into two separate data frames, one with n<30 and one with n>=30
ale1_small <- filter(ale1, n < 30)
ale1_big <- filter(ale1, n >= 30)
ale2_small <- filter(ale2, n < 30)
ale2_big <- filter(ale2, n >= 30)

```

### Point Judith Pond

This section will introduce @fig-pj-stations.

```{r}
#| echo: false
#| label: fig-pj-stations
#| fig-cap: Accumulated local effects (ALE) of station on catch probability in Point Judith Pond. The left panel shows the ALE for all years prior to 2020, and the right panel shows the ALE for the last five years.
#| fig-pos: h

## This code chunk plots the ALE of station for Point Judith Pond

# Define the bounding box for Point Judith Pond
pj_bbox <- c(-71.518, 41.375, -71.49, 41.431)

# Get OSM data for waterbodies
pj_map <- opq(bbox = pj_bbox) %>%
  add_osm_feature(key = 'natural', value = 'water') %>%
  osmdata_sf()



# Plot the historical data
p1 <- ggplot() +
  geom_sf(data = pj_map$osm_multipolygons, fill = "gray80", color = "gray50") +
  coord_sf(xlim = c(pj_bbox[1], pj_bbox[3]), ylim = c(pj_bbox[2], pj_bbox[4])) +
  geom_point(data = ale1_small,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 0.5) +
  geom_point(data = ale1_big,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 1) +
  ggtitle("Point Judith Pond", "Pre-2020") +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(-71.51, -71.49)) +
  scale_y_continuous(breaks = c(41.39, 41.41, 41.43)) +
  scale_size_continuous(breaks = c(0.03, 0.06, 0.09, 0.12)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White overall plot background
  ) +
  guides(
    fill = guide_none(),
    color = guide_legend(title = "Effect"),
    size = guide_legend(title = "Probability")
  )

# Plot the new data
p2 <- ggplot() +
  geom_sf(data = pj_map$osm_multipolygons, fill = "gray80", color = "gray50") +
  coord_sf(xlim = c(pj_bbox[1], pj_bbox[3]), ylim = c(pj_bbox[2], pj_bbox[4])) +
  geom_point(data = ale2_small,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 0.5) +
  geom_point(data = ale2_big,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 1) +
  ggtitle("", 
    paste(max(df2$Haul_Year_Winter) - 4, "to", max(df2$Haul_Year_Winter))) +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(-71.51, -71.49)) +
  scale_y_continuous(breaks = c(41.39, 41.41, 41.43)) +
  scale_size_continuous(breaks = c(0.03, 0.06, 0.09, 0.12)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White overall plot background
  ) +
  guides(
    fill = guide_none(),
    color = guide_legend(title = "Effect"),
    size = guide_legend(title = "Probability")
  )

# Plot Point Judith
ggarrange(p1, p2, ncol = 2, nrow = 1, common.legend = TRUE, legend = "right")

```

### Potter Pond

This section will introduce @fig-pp-stations.

```{r}
#| echo: false
#| label: fig-pp-stations
#| fig-cap: Accumulated local effects (ALE) of station on catch probability in Potter Pond. The left panel shows the ALE for all years prior to 2020, and the right panel shows the ALE for the last five years.
#| fig-pos: h

## This code chunk plots the ALE of station for Potter Pond

# Define the bounding box for Potter Pond
pp_bbox <- c(-71.545, 41.374, -71.5265, 41.40)

# Get OSM data for waterbodies
pp_map <- opq(bbox = pp_bbox) %>%
  add_osm_feature(key = 'natural', value = 'water') %>%
  osmdata_sf()

# Plot the historical baseline
p1 <- ggplot() +
  geom_sf(data = pp_map$osm_multipolygons, fill = "gray80", color = "gray50") +
  coord_sf(xlim = c(pp_bbox[1], pp_bbox[3]), ylim = c(pp_bbox[2], pp_bbox[4])) +
  geom_point(data = ale1_small,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 0.5) +
  geom_point(data = ale1_big,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 1) +
  ggtitle("Potter Pond", "Pre-2020") +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(-71.54, -71.53)) +
  scale_y_continuous(breaks = c(41.37, 41.38, 41.39, 41.4)) +
  scale_size_continuous(breaks = c(0.03, 0.06, 0.09, 0.12)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White overall plot background
  ) +
  guides(
    fill = guide_none(),
    color = guide_legend(title = "Effect"),
    size = guide_legend(title = "Probability")
  )


# Plot the new data
p2 <- ggplot() +
  geom_sf(data = pp_map$osm_multipolygons, fill = "gray80", color = "gray50") +
  coord_sf(xlim = c(pp_bbox[1], pp_bbox[3]), ylim = c(pp_bbox[2], pp_bbox[4])) +
  geom_point(data = ale2_small,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 0.5) +
  geom_point(data = ale2_big,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 1) +
  ggtitle("",
    paste(max(df2$Haul_Year_Winter) - 4, "to", max(df2$Haul_Year_Winter))) +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(-71.54, -71.53)) +
  scale_y_continuous(breaks = c(41.37, 41.38, 41.39, 41.4)) +
  scale_size_continuous(breaks = c(0.03, 0.06, 0.09, 0.12)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White overall plot background
  ) +
  guides(
    fill = guide_none(),
    color = guide_legend(title = "Effect"),
    size = guide_legend(title = "Probability")
  )

# Plot Potter Pond
ggarrange(p1, p2, ncol = 2, nrow = 1, common.legend = TRUE, legend = "right")


```

### Ninigret Pond

This section will introduce @fig-np-stations.

```{r}
#| echo: false
#| label: fig-np-stations
#| fig-cap: Accumulated local effects (ALE) of station on catch probability in Ninigret Pond. The left panel shows the ALE for all years prior to 2020, and the right panel shows the ALE for the last five years.
#| fig-pos: h

## This code chunk plots the ALE of station for Ninigret Pond

# Define the bounding box for Ninigret Pond
np_bbox <- c(-71.7, 41.34, -71.626912, 41.385)

# Get OSM data for waterbodies
np_map <- opq(bbox = np_bbox) %>%
  add_osm_feature(key = 'natural', value = 'water') %>%
  osmdata_sf()

# Plot the historical map
p1 <- ggplot() +
  geom_sf(data = np_map$osm_multipolygons, fill = "gray80", color = "gray50") +
  coord_sf(xlim = c(np_bbox[1], np_bbox[3]), ylim = c(np_bbox[2], np_bbox[4])) +
  geom_point(data = ale1_small, 
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 0.5) +
  geom_point(data = ale1_big,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 1) +
  ggtitle("Ninigret Pond", "Pre-2020") +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(-71.7, -71.67, -71.64)) +
  scale_y_continuous(breaks = c(41.34, 41.36, 41.38)) +
  scale_size_continuous(breaks = c(0.03, 0.06, 0.09, 0.12)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White overall plot background
  ) +
  guides(
    fill = guide_none(),
    color = guide_legend(title = "Effect"),
    size = guide_legend(title = "Probability")
  )

# Plot the new map
p2 <- ggplot() +
  geom_sf(data = np_map$osm_multipolygons, fill = "gray80", color = "gray50") +
  coord_sf(xlim = c(np_bbox[1], np_bbox[3]), ylim = c(np_bbox[2], np_bbox[4])) +
  geom_point(data = ale2_small, 
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 0.5) +
  geom_point(data = ale2_big,
    aes(x = Longitude, y = Latitude, color = posneg, size = abs(0.5 - Probability)),
    alpha = 1) +
  ggtitle("",
    paste(max(df2$Haul_Year_Winter) - 4, "to", max(df2$Haul_Year_Winter))) +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(-71.7, -71.67, -71.64)) +
  scale_y_continuous(breaks = c(41.34, 41.36, 41.38)) +
  scale_size_continuous(breaks = c(0.03, 0.06, 0.09, 0.12)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "white"),  # White panel background
    plot.background = element_rect(fill = "white", color = "white")   # White overall plot background
  ) +
  guides(
    fill = guide_none(),
    color = guide_legend(title = "Effect"),
    size = guide_legend(title = "Probability")
  )

# Plot Ninigret
ggarrange(p1, p2, ncol = 2, nrow = 1, common.legend = TRUE, legend = "right")

```
